---
title: "R Notebook"
author: "Emmanuel"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## Load libraries

```{r}
pacman::p_load(pacman,tidyverse,lubridate,rio)
```

```{r echo=FALSE, message=FALSE,results='hide'}
getwd()
setwd('C:/Users/dell/Desktop/Code/R/Cyclistic')
```

## Import datasets

```{r echo=FALSE, message=FALSE}
jan_21 <- import("2021-divvy-tripdata/202101-divvy-tripdata.csv")
feb_21 <- import("2021-divvy-tripdata/202102-divvy-tripdata.csv")
mar_21 <- import("2021-divvy-tripdata/202103-divvy-tripdata.csv")
apr_21 <- import("2021-divvy-tripdata/202104-divvy-tripdata.csv")
may_21 <- import("2021-divvy-tripdata/202105-divvy-tripdata.csv")
jun_21 <- import("2021-divvy-tripdata/202106-divvy-tripdata.csv")
jul_21 <- import("2021-divvy-tripdata/202107-divvy-tripdata.csv")
aug_21 <- import("2021-divvy-tripdata/202108-divvy-tripdata.csv")
sep_21 <- import("2021-divvy-tripdata/202109-divvy-tripdata.csv")
oct_21 <- import("2021-divvy-tripdata/202110-divvy-tripdata.csv")
nov_21 <- import("2021-divvy-tripdata/202111-divvy-tripdata.csv")
dec_21 <- import("2021-divvy-tripdata/202112-divvy-tripdata.csv")

```

```{r}
colnames(jan_21)
colnames(feb_21)
colnames(mar_21)
colnames(apr_21)
colnames(may_21)
colnames(jan_21) == colnames(feb_21)
colnames(may_21) == colnames(dec_21)






```


### Inspect data for inconsistencies

```{r}
str(jan_21)
str(may_21)
str(nov_21)
str(aug_21)

```

### Stack individual quarter's data frames into one big data frame

```{r echo=FALSE, message=FALSE,results='hide'}
all_trips <- bind_rows( jan_21,feb_21,
                        mar_21,apr_21,
                        may_21,jun_21,
                        jul_21,aug_21,
                        sep_21,oct_21,
                        nov_21,dec_21
                        )
```
## CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS

```{r echo=FALSE, message=FALSE }
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(all_trips)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics
```

### Check member_casuals

```{r}
unique(all_trips$member_casual)
```

### Calculate trip duration (in seconds)

```{r echo=FALSE, message=FALSE}
all_trips$trip_duration <- difftime(all_trips$ended_at,all_trips$started_at,units=c('secs'))
```

### Create day, month, year, fields 
```{r echo=FALSE, message=FALSE,results='hide'}

all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

```


```{r}
str(all_trips)
summary(all_trips)
```

### Drop bad data

```{r echo=-3}
all_trips %>% filter(all_trips$trip_duration<0)
all_trips [all_trips$start_station_name == "HQ QR"]

#all_trips <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$trip_duration<0),]
```

No HQ QR checks recorded in time period


```{r}
all_trips$end_lat %>% is.na() %>% sum()

```



## DESCRIPTIVE ANALYSIS


```{r}
# Compare members and casual users
aggregate(all_trips$trip_duration ~ all_trips$member_casual, FUN = mean)
aggregate(all_trips$trip_duration ~ all_trips$member_casual, FUN = median)
aggregate(all_trips$trip_duration ~ all_trips$member_casual, FUN = max)
aggregate(all_trips$trip_duration ~ all_trips$member_casual, FUN = min)

```

See the average ride time by each day for members vs casual users

```{r}
all_trips$day_of_week <- ordered(all_trips$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

all_trips %>% 
  group_by(all_trips$member_casual, all_trips$day_of_week) %>% 
  summarise(average_trip_duration = mean(trip_duration))
```

```{r}
aggregate(all_trips$trip_duration ~ all_trips$member_casual + all_trips$day_of_week, FUN = mean)
```


2021 Observations
-   Casual users ride have a longer trip duration than members on average
-   Decrease in average trip duration across usertypes with casyual members having up to a 30% decrease in trip durations from 2019
-   Usage pattern similar accross usertypes with weekend spikes in usage unlike in 2019.

### Analyze ridership data by type and weekday

```{r}
all_trips %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%                       #groups by member_casual and weekday
  summarise(number_of_rides = n()                            #calculates the number of rides and average duration 
  ,average_duration = mean(trip_duration)) %>%         # calculates the average duration
  arrange(member_casual, weekday)                                # sorts
```

### Let's visualize the number of rides by rider type

```{r, echo=FALSE, message=FALSE}
all_trips %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(trip_duration)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

### Let's create a visualization for average duration

```{r, echo=FALSE, message=FALSE}
all_trips %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(trip_duration)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

```{r, echo=FALSE, message=FALSE}
all_trips %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(average_duration = mean(trip_duration)) %>%
  ggplot(aes(x = day_of_week, y = average_duration,color=member_casual)) +
  geom_point()+geom_line()
```

## Exports
```{r}
export(all_trips,'all_trips_2021.csv')

```



